---
title: "CPTR330 - Titanic Final Project"
author: Preston Carman
date: June 1, 2017
course: CPTR330
output: html_notebook
---


# Titanic Final Project 

> Homework Notes: Describe the algorithm and give a few of the strengths and weaknesses.

---


```{r}
train <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/train.csv", stringsAsFactors = FALSE)
test <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test.csv", stringsAsFactors = FALSE)
test_final <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test_final.csv", stringsAsFactors = FALSE)

test_results <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test_results.csv", stringsAsFactors = FALSE)
test_results_final <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test_results_final.csv", stringsAsFactors = FALSE)
```

## Step 1 - Collect Data

> Homework Notes: Give an overview of the data and its source.

---



## Step 2 - Exploring And Preparing The Data

> Homework Notes: Explain the data features and any tranformations.

---


### Step 3 - Training A Model On The Data

> Homework Notes: Explain how to train the model.

---



## Step 4 - Evaluating Model Performance

> Homework Notes: Explain the model's perfromance. Highlight key results.

---



## Step 5 - Improving Model Performance

> Homework Notes: What options can be used to improve the model? Explain and show.

---


```{r}
# Kaggle Score
accuracy_kaggle <- 0
accuracy_kaggle
```

```{r}
# Final Score
accuracy_final <- 0
accuracy_final
```
