---
title: "CPTR330 -- Final Project"
author: "Ryan Rabello"
date: "May 31, 2017"
output:
  pdf_document:
    number_section: yes
  html_document: default
course: CPTR330
---

# Titanic Data

## Importing data.
First things first we need to import our data to take a look at it. 
```{r}
get <- function(fileName) {
  # If the file is sucessfully found on the server download it, otherwise use a local copy of the files.
  file <- tryCatch({
    return(read.csv(paste0("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/", fileName), stringsAsFactors = TRUE))
  }, error = function(e) {
    return(read.csv(paste0("./data/titanic/", fileName), stringsAsFactors = TRUE))
  })
  return(file)
}
raw_train <- get("train.csv")
raw_test <- get("test.csv")
raw_test_final <- get("test_final.csv")
raw_test_results <- get("test_results.csv")
raw_test_results_final <- get("test_results_final.csv")

# For testing purposes. 
#train <- read.csv("./data/train.csv", stringsAsFactors = TRUE)
#test_final <- read.csv("./data/test.csv", stringsAsFactors = TRUE)
str(raw_train)
```

## Explaining data

The following list briefly explains what each variable is and what type (categorical or regression) it is.

* **Survived**	A Boolean (0 or 1) indicating the survival of this particular passenger. 
* **pclass**	Ticket class	1 = 1st, 2 = 2nd, 3 = 3rd (A numerical representation of the)
* **sex**	Gender of the individual. 
* **Age**	Age in floating point years.
* **sibsp**	# of siblings / spouses aboard the Titanic
* **parch**	# of parents / children aboard the Titanic	
* **ticket**	Ticket number
* **fare**	Passenger fare (the cost of the ticket).
* **cabin**	Cabin number (String ex "A15" and "B12")
* **embarked**	Port of Embarkation	C = Cherbourg, Q = Queenstown, S = Southampton

Because `PassengerId`, `Name` and `ticket` are mostly unique we they will not be useful in any machine learning calculation and are therefore nullified. I'm also changing some of the names of variables so that everything works okay. Lastly I'm simplifying the cabin variable to just the first letter of the cabin. I believe this corresponds to the floor of the cabin. 
```{r}
process <- function (titanic) {
  titanic$PassengerId <- NULL
  titanic$Name <- NULL
  titanic$Ticket <- NULL
  levels(titanic$Embarked) <- c("Missing", "Cherbourg", "Queenstown",  "Southampton")
  titanic$Pclass <- as.factor(titanic$Pclass)
  if ("Survived" %in% names(titanic)){
    titanic$Survived <- factor(titanic$Survived, c(0, 1), c("Died", "Survived"))
  }
  
  # Convert cabin to "floor" (the first letter of each cabin)
  cabinLev <- levels(titanic$Cabin)
  cabinLev <- substr(cabinLev, 1,1)
  levels(titanic$Cabin) <- cabinLev
  levels(titanic$Cabin) <- c("?","A", "B", "C", "D", "E", "F", "G", "T")
  titanic$Floor <- titanic$Cabin
  
  
  # Convert Cabin to `Room`
  titanic$Room <- as.numeric(substring(as.character(titanic$Cabin), 2))
  
  titanic$Cabin <- NULL
  
  return(titanic)
}
train <- process(raw_train)
test <- process(raw_test)
test_final <- process(raw_test_final)
test_results_final <- factor(raw_test_results_final$Survived, c(0, 1), c("Died", "Survived"))

test_all <- cbind(Survived = test_results_final, test_final)

titanic <- rbind(train, test_all)

```

A global function for exporting data. `data` is a vector of answers and `name` the name of the csv that will be exported. (ex. name = "tree" -> "tree.csv")
```{r}
export <- function(data,name) {
  kaggle <- data.frame(PassengerId = raw_test_final$PassengerId, Survived = as.numeric(data) - 1)
  write.csv(x = kaggle, file = paste("./exports/",  name, ".csv", sep = ""), row.names = F, quote = F)
}
```

## Analyzing some data. 

```{r}
# Dynamically load `ggplot2`
package <- 'ggplot2'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}
package <- 'ggthemes'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}


#ggplot <- function(...) ggplot2::ggplot(...) + scale_color_fivethirtyeight("cyl") + theme_fivethirtyeight()
# ggplot <- function(...) ggplot2::ggplot(...) + theme_hc() + scale_colour_hc()
ggplot(titanic, aes(Survived, Fare)) + geom_violin(scale = "area") + ggtitle("Fare related to survival")
ggplot(titanic, aes(Survived, Age)) + geom_violin(scale = "area") + ggtitle("Age related to survival")

# Plot a jitter graph
ggplot(titanic, aes(Pclass, Survived)) + geom_jitter() + ggtitle("Person Class vs Survival")
ggplot(titanic, aes(Sex, Survived)) + geom_jitter() + ggtitle("Gender vs Survival")
ggplot(titanic, aes(x = Floor, y = Room, colour = Survived)) + geom_jitter() + ggtitle("Cabin vs Survival")
ggplot(titanic, aes(x=SibSp, y=Parch, colour = Survived)) + geom_jitter() + ggtitle("Family vs Survival") +labs(x="Siblings/Spouse", y="Parents/Children")
```

# Descision Trees

## Training
```{r}
# Dynamically load `C50`
package <- 'C50'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}

# Preprocess
names <- levels(train$Cabin)
names[1] <- "?"
levels(train$Cabin) <- names
train$Cabin = train$Cabin

# Train the model. 
tree <- C5.0(train[,-1], train$`Survived`, trials = 5)

summary(tree)
# plot(tree)
```


## Testing
```{r}
tree_pred <- predict(tree, test)

table(tree_pred, test_results_final)

# Calculate the percentage of correct guesses.
tree_correct <- tree_pred == test_results_final
table(tree_correct)/length(tree_correct) * 100
```

## Test 2
```{r}
tree <- C5.0(train[,-1], train$`Survived`, trials = 10, control = C5.0Control(bands = 50), rules = T)

# summary(tree)

# Testing
tree_pred <- predict(tree, test)

# Calculate the percentage of correct guesses.
tree_correct <- tree_pred == test_results_final
table(tree_correct)/length(tree_correct) * 100
```

## Kaggle Exporting
```{r}
tree_pred <- predict(tree, test_final)
export(tree_pred, "tree")
```

# Naive Bayes

```{r}
# Load e1071 if it's not installed install it. 
package <- 'e1071'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}
# Trains the data set.
nb1 <- naiveBayes(Survived~., data=train)
#nb1
```

## Testing 
```{r}
nb_guess <- predict(nb1, test[,c("Pclass", "Sex", "Age", "SibSp", "Parch", "Fare", "Cabin")])
table(nb_guess, test_results_final)

# Calculate the percentage of correct guesses.
nb_correct <- nb_guess == test_results_final
table(nb_correct)/length(nb_correct) * 100

export(nb_guess, "nb")
```

# Neural Networks

Neural networks only work with numerical data so I'm going to for now, remove the categorical data. An option to do later would be to associate a category to a set of nodes and activate the corresponding input node for the specific category.
```{r}
nn_process <- function(titanic){
  
  for(key in c("Cabin", "Embarked", "Pclass")) {
    # Convert those factors into boolean data.frames
    a <- NULL
    for(level in levels(titanic[[key]])) {
      a[[level]] <- as.numeric(titanic[key] == level)
    }
    # Add cbind all those data.frames together
    titanic <- cbind(titanic, as.data.frame(a))
    titanic[key] <- NULL
  }
  
  
  # Covert Everything to a numeric.
  titanic <- lapply(titanic, function(x) as.numeric(x))
  
  # Define our Normalize function
  normalize <- function(x) {
    if (anyNA(x) || max(x) == min(x)) {
      x <- scale(x, center = FALSE, scale = TRUE)
      x[is.na(x)] <- 0
      return(x)
    } else {
      return ((x - min(x)) / (max(x) - min(x)))
    }
  }
  
  titanic <- lapply(titanic, normalize)
  titanic <- data.frame(titanic)
  return(titanic)
}

nn_train <- nn_process(train)
nn_test <- nn_process(test)
```
Note that the cbind.all function() was found [here](https://stackoverflow.com/questions/26684072/add-columns-to-an-empty-data-frame-in-r).

## Training the neural network
```{r}
# Dynamically load `neuralnet`
package <- 'neuralnet'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}

set.seed(1234)

# Create a formula in the form "Survived ~ V(1) + V(2) + ... + V(n-1) + V(n)"
fmla <- as.formula(paste("Survived ~ ", paste(names(nn_test), collapse= "+")))

# Train the model
nn <- neuralnet(fmla , data = nn_train, hidden = c(7, 7), threshold = 0.05)
plot(nn)
```

```{r}
# Predict some data 
nn_predict <- compute(nn, subset(nn_train, select=-c(Survived)))
cor(nn_predict$net.result, as.numeric(nn_train$Survived) - 1)
```

## Process and export data
```{r}
nn_predict <- compute(nn, nn_test)
results <- round(nn_predict$net.result)
results[results < 0] <- 0
results[results > 1] <- 1
results <- factor(x = results, c(0,1), c("Dead", "Survived"))

export(results, "nn")
```

