---
title: "CPTR330 -- Final Project"
author: "Ryan Rabello"
date: "May 31, 2017"
output:
  pdf_document:
    number_section: yes
  html_document: default
course: CPTR330
---

# Titanic Data

## Importing data.
First things first we need to import our data to take a look at it. 
```{r}
train <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/train.csv", stringsAsFactors = TRUE)
test <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test.csv", stringsAsFactors = TRUE)
test_final <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test_final.csv", stringsAsFactors = TRUE)
test_results <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test_results.csv", stringsAsFactors = TRUE)
test_results_final <- read.csv("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/test_results_final.csv", stringsAsFactors = TRUE)

# For testing purposes. 
train <- read.csv("./data/train.csv", stringsAsFactors = TRUE)
test_final <- read.csv("./data/test.csv", stringsAsFactors = TRUE)
str(train)
```

## Explaining data

The following list briefly explains what each variable is and what type (categorical or regression) it is.

* **Survived**	A Boolean (0 or 1) indicating the survival of this particular passenger. 
* **pclass**	Ticket class	1 = 1st, 2 = 2nd, 3 = 3rd (A numerical representation of the)
* **sex**	Gender of the individule. 
* **Age**	Age in floating point years.
* **sibsp**	# of siblings / spouses aboard the Titanic
* **parch**	# of parents / children aboard the Titanic	
* **ticket**	Ticket number
* **fare**	Passenger fare (the cost of the ticket).
* **cabin**	Cabin number (String ex "A15" and "B12")
* **embarked**	Port of Embarkation	C = Cherbourg, Q = Queenstown, S = Southampton

Because `PassengerId`, `Name` and `ticket` are mostly unique we they will not be useful in any machine learning calculation and are therefore nullified. I'm also changing some of the names of vairables so that everything works okay. Lastly I'm simplifying the cabin variable to just the first letter of the cabin. I believe this corresponds to the floor of the cabin. 
```{r}
process <- function (titanic) {
  if ("Survived" %in% names(titanic)){
    titanic$Survived <- factor(titanic$Survived, c(0, 1), c("Died", "Survived"))
  }
  titanic$PassengerId <- NULL
  titanic$Name <- NULL
  titanic$Ticket <- NULL
  levels(titanic$Embarked) <- c("Missing", "Cherbourg", "Queenstown",  "Southampton")
  
  # Plot only cabins with more than 3 people. 
  cabinLev <- levels(titanic$Cabin)
  cabinLev <- substr(cabinLev, 1,1)
  levels(titanic$Cabin) <- cabinLev
  levels(titanic$Cabin) <- c("?","A", "B", "C", "D", "E", "F", "G", "T")
  return(titanic)
}
train_titanic <- process(train)
test_titanic <- process(test_final)
test_titanic_labels <- factor(test_results_final$Survived, c(0, 1), c("Died", "Survived"))

test_titanic_all <- cbind(Survived = test_titanic_labels, test_titanic)

titanic <- rbind(train_titanic, test_titanic_all)

```

## Analyzing some data. 

```{r}
# Dynamically load `ggplot2`
package <- 'ggplot2'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}
package <- 'ggthemes'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}


#ggplot <- function(...) ggplot2::ggplot(...) + scale_color_fivethirtyeight("cyl") + theme_fivethirtyeight()
# ggplot <- function(...) ggplot2::ggplot(...) + theme_hc() + scale_colour_hc()
ggplot(titanic, aes(Survived, Fare)) + geom_violin(scale = "area") + ggtitle("Fare related to survival")
ggplot(titanic, aes(Survived, Age)) + geom_violin(scale = "area") + ggtitle("Age related to survival")

# Plot a jitter graph
ggplot(titanic, aes(Pclass, Survived)) + geom_jitter() + ggtitle("Person Class vs Survival")
ggplot(titanic, aes(Sex, Survived)) + geom_jitter() + ggtitle("Gender vs Survival")
ggplot(titanic, aes(Cabin, Survived)) + geom_jitter() + ggtitle("Cabin Level vs Survival")
ggplot(titanic, aes(x=SibSp, y=Parch, colour = Survived)) + geom_jitter() + ggtitle("Family vs Survival") +labs(x="Siblings/Spouse", y="Parents/Children")
```

# Descision Trees

## Training
```{r}
# Dynamically load `C50`
package <- 'C50'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}

# Preprocess
names <- levels(train$Cabin)
names[1] <- "?"
levels(train$Cabin) <- names
train_titanic$Cabin = train$Cabin

# Train the model. 
tree <- C5.0(train_titanic[,-1], train_titanic$`Survived`, trials = 5)

summary(tree)
# plot(tree)
```


## Testing
```{r}
tree_pred <- predict(tree, test_titanic)

table(tree_pred, test_titanic_labels)

# Calculate the percentage of correct guesses.
tree_correct <- tree_pred == test_titanic_labels
table(tree_correct)/length(tree_correct) * 100
```

## Test 2
```{r}
tree <- C5.0(train_titanic[,-1], train_titanic$`Survived`, trials = 10, control = C5.0Control(bands = 50), rules = T)

summary(tree)
```
## Kaggle Exporting
```{r}
export <- function(data,name) {
  kaggle <- data.frame(PassengerId = test_final$PassengerId, Survived = as.numeric(data) - 1)
  write.csv(x = kaggle, file = paste("./exports/",  name, ".csv", sep = ""), row.names = F, quote = F)
}

export(tree_pred, "tree")
```

# Naive Bayes

```{r}
# Load e1071 if it's not installed install it. 
package <- 'e1071'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}
# Trains the data set.
nb1 <- naiveBayes(Survived~., data=train_titanic)
#nb1
```

## Testing 
```{r}
nb_guess <- predict(nb1, test_titanic[,c("Pclass", "Sex", "Age", "SibSp", "Parch", "Fare", "Cabin")])
table(nb_guess, test_titanic_labels)

# Calculate the percentage of correct guesses.
nb_correct <- nb_guess == test_titanic_labels
table(nb_correct)/length(nb_correct) * 100

export(nb_guess, "nb")
```

# Neural Networks

Neural networks only work with numerical data so I'm going to for now, remove the categorical data. An option to do later would be to associate a category to a set of nodes and activate the corresponding inpute node for the specific category.
```{r}
nn_process <- function(titanic){
  # Covert Everything to a numeric.
  titanic <- lapply(titanic, function(x) as.numeric(x))
  
  # Define our Normalize function
  normalize <- function(x) {
    if (anyNA(x)) {
      x <- scale(x, center = FALSE, scale = TRUE)
      x[is.na(x)] <- 0
      return(x)
    } else {
      return ((x - min(x)) / (max(x) - min(x)))
    }
  }
  
  titanic <- lapply(titanic, normalize)
  titanic <- data.frame(titanic)
  return(titanic)
}

nn_train <- nn_process(train_titanic)
nn_test <- nn_process(test_titanic)
```

## Training the neural network
```{r}
# Dynamically load `neuralnet`
package <- 'neuralnet'
if(package %in% rownames(installed.packages())) {
  do.call('library', list(package))
} else {
  install.packages(package)
  do.call("library", list(package))
}

set.seed(1234)

# Create a formula in the form "Survived ~ V1 + V2 + ... + V33 + V34"
fmla <- as.formula(paste("Survived ~ ", paste(names(nn_test), collapse= "+")))

# Train the model
nn <- neuralnet(fmla , data = nn_train, hidden = c(3))
plot(nn)
```

```{r}
# Predict some data 
nn_predict <- compute(nn, subset(nn_train, select=-c(Survived)))
cor(nn_predict$net.result, as.numeric(nn_train$Survived) - 1)
```

## Process and export data
```{r}
nn_predict <- compute(nn, nn_test)
results <- round(nn_predict$net.result)
results[results < 0] <- 0
results[results > 1] <- 1
results <- factor(x = results, c(0,1), c("Dead", "Survived"))

export(results, "nn")
```


# knn?

# 
