---
title: "CPTR330 - Titanic Final Project"
author: Preston Carman
date: June 1, 2017
course: CPTR330
output: html_notebook
---


# Titanic Final Project 

> Homework Notes: Describe the algorithm and give a few of the strengths and weaknesses.

---


## Step 1 - Collect Data

> Homework Notes: Give an overview of the data and its source.

---


```{r}
# This function gets the target csv file from the url. If that fails it uses a copy of the files stored locally. 
get <- function(fileName) {
  
  # return(read.csv(paste0("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/", fileName), stringsAsFactors = FALSE))
  
  # If the file is sucessfully found on the server download it, otherwise use a local copy of the files.
  file <- tryCatch({
   return(read.csv(paste0("https://cs.wallawalla.edu/~carmpr/cptr330/titanic/", fileName), stringsAsFactors = F))
  }, error = function(e) {
   return(read.csv(paste0("./data/titanic/", fileName), stringsAsFactors = TRUE))
  })
  return(file)
}
# The train.csv contains all features and is used to train models. 
train <- get("train.csv")

# The test.csv is not going to change during the final time. 
test <- get("test.csv")

# The test_final.csv is what is going to be used (like a kaggle submission) to grade the performance of our algorithm. 
test_final <- get("test_final.csv")

# The test_results.csv contains the labels (or answers) to test.csv.
test_results <- get("test_results.csv")

# The test_results_final.csv contains the labels for test_final.csv.
test_results_final <- get("test_results_final.csv")
```


## Step 2 - Exploring And Preparing The Data

> Homework Notes: Explain the data features and any tranformations.

---


```{r}
train$Survived <- as.factor(train$Survived)
train$Pclass <- as.factor(train$Pclass)

test$Pclass <- as.factor(test$Pclass)
test_results$Survived <- as.factor(test_results$Survived)

test_final$Pclass <- as.factor(test_final$Pclass)
test_results_final$Survived <- as.factor(test_results_final$Survived)
```

### Step 3 - Training A Model On The Data

> Homework Notes: Explain how to train the model.

---


```{r}
library(e1071)
nb <- naiveBayes(Survived~Sex + Pclass, data=train, laplace = 1)
```


## Step 4 - Evaluating Model Performance

> Homework Notes: Explain the model's perfromance. Highlight key results.

---



## Step 5 - Improving Model Performance

> Homework Notes: What options can be used to improve the model? Explain and show.

---


```{r}
# Kaggle Score
prediction <- predict(nb, test, type="class")
xtab <- table(prediction, test_results$Survived)
xtab

accuracy_kaggle <- sum(diag(xtab))/sum(xtab)
accuracy_kaggle
```

```{r}
# Final Score
prediction_final <- predict(nb, test_final, type="class")
xtab <- table(prediction_final, test_results_final$Survived)
xtab

accuracy_final <- sum(diag(xtab))/sum(xtab)
accuracy_final
```
